---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
```{r load_libraries, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
```

The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day. You can download the data here: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip)

## Loading and preprocessing the data
Read data from csv file, and format the date fields as Date type.
```{r load_prep_data}
data <- read.csv("activity.csv", header = T)
data$date <- as.Date(data$date)
summary(data$steps)
```

## What is mean total number of steps taken per day?
Summarise the total number of steps taken per day and calculate some statistics.
```{r total_steps_per_day}
day_steps <- data %>% group_by(date) %>% summarise(total=sum(steps, na.rm = T))
```

The mean is: ```r mean(day_steps$total, na.rm=T)```
The median is: ```r median(day_steps$total, na.rm=T)```
```{r plot_total_steps_per_day, message=FALSE, warning=FALSE}
ggplot(day_steps, aes(x=total)) +
  geom_histogram(color = "black", fill="gray") +
  labs(x="Number of steps", title="Total number of steps taken each day") +
  geom_vline(aes(xintercept=median(day_steps$total), color="median")) +
  geom_vline(aes(xintercept=mean(day_steps$total), color="mean")) + 
  theme(legend.position = "bottom")
```

## What is the average daily activity pattern?
Summarise the average number of steps taken in every 5-minute interval for each day.
```{r avg_steps}
interval_steps <- data %>% group_by(interval) %>% summarise(avg=mean(steps, na.rm = T))
```

The 5-minute interval ```r interval_steps[which(interval_steps$avg==max(interval_steps$avg)),"interval"]``` on average accross all the days in the dataset, contains the maximum number of steps
```{r plot_avg_steps}
ggplot(interval_steps, aes(x=interval, y=avg)) +
  geom_line() +
  labs(x="Interval", y="Average of steps", title="Average number of steps")
```

## Imputing missing values
There are some missing values on data. The total missing rows are: ```r table(is.na(data))[2][[1]]```

Those missing values can introduce bias into some calculations, the mean of the interval can be used to fill those gaps.
```{r fill_nas}
fill.na <- function(x){
  replace(x, is.na(x), mean(x, na.rm = T))
}
data_complete <- data %>% group_by(interval) %>% mutate(steps=fill.na(steps))
day_steps_complete <- data_complete %>% group_by(date) %>% summarise(total=sum(steps))
```

The mean is


- Filling NA's: ```r mean(day_steps_complete$total)```
- Without filling NA's: ```r mean(day_steps$total, na.rm=T)```


The median is


- Filling NA's: ```r median(day_steps_complete$total)```
- Without filling NA's: ```r median(day_steps$total, na.rm=T)```
```{r plot_total_complete_steps_per_day, message=FALSE, warning=FALSE}
ggplot(day_steps_complete, aes(x=total)) +
  geom_histogram(color = "black", fill="gray") +
  labs(x="Number of steps", title="Total number of steps taken each day") +
  geom_vline(aes(xintercept=median(day_steps_complete$total), color="median")) +
  geom_vline(aes(xintercept=mean(day_steps_complete$total), color="mean")) + 
  theme(legend.position = "bottom")
```

By filling the missing values with the mean of the interval, the overall mean of the dataset steps increased. The same happens to the median.

## Are there differences in activity patterns between weekdays and weekends?
To calculate if a date corresponds to weekday or weekend the function weekdays() is going to be used.
```{r day_type}
data <- data %>% mutate(day.type=ifelse(grepl("^S",weekdays(date)), "weekend", "weekday"))
data$day.type <- as.factor(data$day.type)
interval_daytype_steps <- data %>% group_by(interval, day.type) %>% summarise(avg=mean(steps, na.rm = T))
```

In weekdays the average of steps is higher in the 5-minute intervals from 500 to 1000. In the weekend the average of steps increases by the end of the 5-minute intervals, in comparisson to weekdays.
```{r plot_day_type}
ggplot(interval_daytype_steps, aes(x=interval, y=avg)) +
  geom_line() +
  labs(x="Interval", y="Average of steps", title="Average number of steps by Day type") +
  facet_grid(cols = vars(day.type))
```

